Unifying intra- and inter-specific variation in tropical tree mortality

### 2017-11-27 ms/manuscript.Rnw
updated file names

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a small, but significant, increase in predictive skill (Fig. \ref{fig2}A).

@@

\subsection*{Hazard functions}
Comparing the three hazard functions in Fig. \ref{fig1}B, we found that the third function --- with both growth-dependent and growth-independent terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only or growth-dependent only functions (Fig. \ref{fig2}A). Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig. \ref{fig2}A). A summary of hyper parameter estimates can be found in Table \ref{table:1}.
<<param-table,results="asis",center=TRUE, echo= FALSE>>= 
param_table(final_model)
@
\begin{figure}[h!]
\includegraphics{figures/fig3.pdf}
\caption{Relationship between species growth-dependent mortality rate and wood density. Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
\label{fig3}
\end{figure}

\subsection*{Wood density and other species effects}
Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig. \ref{fig2} A,B). The most parsimonious model, with highest predictive skill, was that attributing the wood density effect to the growth-independent hazard term ($\gamma$; Fig. \ref{fig2} B). Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate, $\gamma$ (Fig. \ref{fig3}). This meant that fast growing individuals from a low wood density species had, on average, higher mortality rates (Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species. For example, a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$. Incorporating an additional species-level random effect to capture any additional inter specific differences substantially improved model skill relative to a model with only wood density (Fig. \ref{fig2} A).

\begin{figure}[t]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in instantaneous mortality rates (A,B), and annual mortality probability (C,D), with increasing growth rates. A,C) Estimated species curves for all 203 BCI species used in this study. B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species. Note: Species parameter estimates can be found in the supplementary Figures S3-S5}
\label{fig4}
\end{figure}

\subsection*{Proportion of variation explained}
To compliment the above analysis based on skill in predicting held-out data, we used the final fitted model to estimate the amount of variation being contributed by different factors in 1-yr survival for all individuals (Fig. \ref{fig5}). All species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr survival. However, wood density alone captured only \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of total variation. Census effects accounted for an even smaller fraction: only \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation. Comparing the two independent hazards, growth-dependent hazards (i.e. the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for a smaller proportion (Fig. \ref{fig5}).

@@

\begin{figure}[t]
\includegraphics{figures/fig5.pdf}
\caption{Proportional reduction in predicted 1-year survival across all individuals from removing different effects. Effects are not mutually exclusive so sum to more than 1.0. Note also that the effect of "wood density" is a subset of the overall "species" effect.}  
\label{fig5}
\end{figure}

>>>>

\subsection*{Mortality over time}
Comparing the three 5-year intervals between censuses from 1995 to 2010, we found average mortality rates progressively decreasing over time. The highest proportion of trees death occurred between 1995\textendash 2000 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 1)}\%) followed by 2000\textendash 2005 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 2)}\%) and 2005\textendash 2010 (\Sexpr{prop_deaths(BCI_model_dataset_true_dbh, 3)}\%). Consequently, when we allowed hazard rates to be scaled by census (i.e. adding term $\delta$) we observed a small, but significant, increase in predictive skill (Fig. \ref{fig2}A). Individual census effects can be found in Table S1.

@@

\subsection*{Hazard functions}
Comparing the three hazard functions in Fig. \ref{fig1}B, we found that the third function --- with both growth-dependent and growth-independent terms, i.e. $(\alpha e^{-\beta X_{i}} + \gamma)\delta_t$ --- significantly outperformed both the growth-independent only or growth-dependent only functions (Fig. \ref{fig2}A). Moreover, we found that predictive skill was higher when using stem diameter growth over stem-area growth (Fig. \ref{fig2}A). A summary of hyper parameter estimates can be found in Table S1.

\begin{figure}[h!]
\includegraphics{figures/fig3.pdf}
\caption{Relationship between species growth-dependent mortality rate and wood density. Points are estimated mean baseline mortality rates $\pm$ 95\% credible intervals for each of the 203 BCI species used in this study. Blue trendline with grey shading shows average ($\pm$ 95\% credible intervals) expected relationship.}
\label{fig3}
\end{figure}

\subsection*{Wood density and other species effects}
Including wood density as an effect on either growth-independent or growth-dependent parameters significantly improved model skill relative to a model without such effects (Fig. \ref{fig2} A,B). The most parsimonious model, with highest predictive skill, was that attributing the wood density effect to the growth-independent hazard term ($\gamma$; Fig. \ref{fig2} B). Specifically, we found that wood density was negatively correlated with a species' baseline mortality rate, $\gamma$ (Fig. \ref{fig3}). This meant that fast growing individuals from a low wood density species had, on average, higher mortality rates (Fig. \ref{fig4} A,B), and thus higher annual probability of death (Fig. \ref{fig4} C,D), relative to fast growing individuals of high wood density species. For example, a species with a wood density of 0.3 g $\textrm{cm}^{-3}$ had an estimated mean probability of dying of \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.3),2)['mean']} yr$^{-1}$ compared to \Sexpr{round(extract_baseline_estimate(final_model,FALSE,0.8),2)['mean']} yr$^{-1}$ for a species with a wood density of 0.8 g $\textrm{cm}^{-3}$. Incorporating an additional species-level random effect to capture any additional inter specific differences substantially improved model skill relative to a model with only wood density (Fig. \ref{fig2}A).

\begin{figure}[t]
\includegraphics{figures/fig4.pdf}
\caption{Predicted changes in instantaneous mortality rates (A,B), and annual mortality probability (C,D), with increasing growth rates. A,C) Estimated species curves for all 203 BCI species used in this study. B,D) The average ($\pm$ 95\% credible intervals) expected curve for a high (0.8) and low (0.3) wood density species. Note: Species parameter estimates can be found in the supplementary Figures S3-S5.}
\label{fig4}
\end{figure}

\subsection*{Proportion of variation explained}
To compliment the above analysis based on skill in predicting held-out data, we used the final fitted model to estimate the amount of variation captured by different factors in 1-yr survival for all individuals (Fig. \ref{fig5}). Species-level effects contributed ~\Sexpr{extract_prop_explained(param_variance_explained, I('species'))}\% of predicted variation in 1-yr survival. Wood density explained \Sexpr{extract_prop_explained(param_variance_explained, I('wood_density'))}\% of this species variation. Census accounted for only \Sexpr{extract_prop_explained(param_variance_explained, I('census'))}\% of the total predicted variation. The growth-dependent hazard (i.e. the part of the model allowed to vary by individual growth) accounted for \Sexpr{extract_prop_explained(param_variance_explained, I('growth_dependent'))}\% of the total variation, while growth-independent hazards accounted for the remainder (Fig. \ref{fig5}).

@@

\begin{figure}[t]
\includegraphics{figures/fig5.pdf}
\caption{Variation captured in predicted 1-year survival across all individuals as determined by removing different effects. Effects are not mutually exclusive thus sum to more than 1.0. Note "wood density" is a subset of the overall "species" effect.}  
\label{fig5}
\end{figure}
